---
title: "Reproducible Research: Peer Assessment 1"
author: "Ben Holcombe"
date: "Sunday, April 19, 2015"
output: html_document
keep_md: true
---

## Loading and preprocessing the data
- We begin by downloading the [repdata_data_activity.zip][1] file.  
[1]: http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip [repdata_data_activity.zip]
- Once downloaded we unzipping the file.
```{r}
fileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL, destfile = "./activity.zip")
unzip("activity.zip")
```

- Next the file is loaded into a data frame and the date column is converted from a factor class into a date class.
```{r}
ActivityData <- read.csv("activity.csv")
ActivityData$date <- as.Date(ActivityData$date)
```

## What is mean total number of steps taken per day?
- Calculate the total steps for each day.
```{r}
ActivitySum <- aggregate(steps ~ date, FUN = sum, data = ActivityData)
colnames(ActivitySum)[2] <- "steps.total"
```
- Display a histogram of the total number of steps per day.
```{r}
hist(ActivitySum$steps, breaks = 288)
```

- Calculate the Mean and Median steps for each day.
```{r}
ActivityMean <- aggregate(steps ~ date, FUN = mean, data = ActivityData)
colnames(ActivityMean)[2] <- "steps.mean"
ActivityMedian <- aggregate(steps ~ date, FUN = median, data = ActivityData)
colnames(ActivityMedian)[2] <- "steps.median"
ActivitySummary <- cbind(ActivitySum, steps.mean = ActivityMean$steps.mean, steps.median = ActivityMedian$steps.median)
```

The Mean, Median, and Total steps taken per day.
```{r}
ActivitySummary
```
## What is the average daily activity pattern?
- Calculate the total steps for each interval.
```{r}
ActivityIntervalSum <- aggregate(steps ~ interval, FUN = sum, data = ActivityData)
```
- Below is the total steps taken by interval.  This graph shows a peak early on.  
```{r}
ActivityIntervalAvg <- aggregate(steps ~ interval, FUN = mean, data = ActivityData)
library(ggplot2)
ChartTitle <- "Total Steps taken by Interval"
ggplot(data = ActivityIntervalAvg, aes(interval, steps)) + geom_line() + labs(title = ChartTitle)
```

- Calculating the peak interval.
```{r}
PeakInterval <- ActivityIntervalSum[order(-ActivityIntervalSum["steps"]),]
PeakInterval[1,]
pinterval <- PeakInterval[1,1]
psteps <- PeakInterval[1,2]
```

- The peak interval is `r pinterval` with a total of `r psteps` steps taken.

## Imputing missing values
```{r}
ActivityDataMissing <- is.na(ActivityData$steps)
missing <- as.character(t(ActivityDataMissing) %*% ActivityDataMissing)
```
- The total number of missing values is `r missing`.

- When we substitute 1 for the missing values we have the following results.
```{r}
ActivityData[is.na(ActivityData)] <- 1
```

```{r}
ActivitySum <- aggregate(steps ~ date, FUN = sum, data = ActivityData)
colnames(ActivitySum)[2] <- "steps.total"
```
- Display a histogram of the total number of steps per day.
```{r}
hist(ActivitySum$steps, breaks = 288)
```

- Calculate the Mean and Median steps for each day.
```{r}
ActivityMean <- aggregate(steps ~ date, FUN = mean, data = ActivityData)
colnames(ActivityMean)[2] <- "steps.mean"
ActivityMedian <- aggregate(steps ~ date, FUN = median, data = ActivityData)
colnames(ActivityMedian)[2] <- "steps.median"
ActivitySummary <- cbind(ActivitySum, steps.mean = ActivityMean$steps.mean, steps.median = ActivityMedian$steps.median)
```

The Mean, Median, and Total steps taken per day.
```{r}
ActivitySummary
```

Since this is a very simple substitution of values you see the histogram shift where the missing values are represented in the 0 - 5000 range.  

## Are there differences in activity patterns between weekdays and weekends?
```{r}
ActivityData$day <- weekdays(ActivityData$date)
ActivityData$week.part <- sapply(ActivityData$day, switch,
    Saturday = "weekend",
    Sunday = "weekend",
    Monday = "weekday",
    Tuesday = "weekday",
    Wednesday = "weekday",
    Thursday = "weekday",
    Friday = "weekday",
  )

ggplot(data = ActivityData, aes(x=interval, y=steps)) + geom_line() + facet_wrap( ~ week.part, ncol=1)

```